A Step-Indexed Logic in Agda

### 2023-06-19 papers/cpp2024/paper.tex
check

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
\newcommand{\dotspace}{.\,}
\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\dyn}{\star}
\newcommand{\Dyn}{\ensuremath{\dyn}}
\newcommand{\Int}{\key{int}}
\newcommand{\Bool}{\key{bool}}
\newcommand{\lam}[1]{\lambda #1 \dotspace}
\newcommand{\app}{\;}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\inj}[2]{#1 \,!\, #2}
\newcommand{\proj}[2]{#1 \,?\, #2}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Step-indexed logical relations are an important tool in the
  metatheory of programming
  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
  can be streamlined by use of modal logics such as
  LSLR~\citep{Dreyer:2011wl}.  The Iris~\citep{JUNG:2018aa} system
  provides support for reasoning about step-indexed logical relations
  in the Coq proof assistant, but there is not yet analogous support
  in the Agda proof assistant. This paper describes how to embed a
  step-indexed logic in Agda and application of the logic to a
  semantic type safety proof for an internal language for the
  Gradually-Typed Lambda Calculus.
\end{abstract}

@@

\section{Motivation}

Logical relations provide a powerful tool for proving properties of
programming languages, from type safety to relational parametricity,
the gradual guarantee, and noninterference.  Initially, logical
relations were applied to terminating languages such as the Simply
Typed Lambda Calculus~\citep{Tait:1967aa} and System
F~\citep{GIRARD72}, but \citet{Appel:2001aa} extended the technique to
handle languages with nontermination by introducing a step index into
the logical relation. After considerable experience with applying
step-indexed logical
relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
reasoning about step indexes.

>>>>

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
\newcommand{\dotspace}{.\,}
\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\dyn}{\star}
\newcommand{\Dyn}{\ensuremath{\dyn}}
\newcommand{\Int}{\key{int}}
\newcommand{\Bool}{\key{bool}}
\newcommand{\lam}[1]{\lambda #1 \dotspace}
\newcommand{\app}{\,}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\inj}[2]{#1 \,!\, #2}
\newcommand{\proj}[2]{#1 \,?\, #2}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Step-indexed logical relations are an important tool in the
  metatheory of programming
  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
  can be streamlined by use of modal logics such as
  LSLR~\citep{Dreyer:2011wl}.  The Iris system~\citep{JUNG:2018aa}
  provides support for reasoning about step-indexed logical relations
  in the Coq proof assistant, but there is not yet analogous support
  for the Agda proof assistant. This paper describes an embedding of
  step-indexed logic in Agda and an application of the logic to a
  semantic type safety proof for an internal language for the
  Gradually-Typed Lambda Calculus.
\end{abstract}

@@

%TODO: Road Map/Outline of what's in the paper

\section{Introduction}

Logical relations are a useful technique for proving properties of
programming languages, from type safety to relational parametricity,
and noninterference.  Initially, logical relations were applied to
terminating languages such as the Simply Typed Lambda
Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}, but
\citet{Appel:2001aa} extended the technique to handle languages with
nontermination by introducing a step index into the logical
relation. After some experience with applications of step-indexed
logical
relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
reasoning about step indexes. This research program eventually
produced Iris~\citep{JUNG:2018aa}, a system that supports a
combination of step indexed logical relations and separation logic
within the Coq proof
assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.

The author is interested in step-indexed logical relations for
purposes of developing mechanized proofs in Agda of the Gradual
Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
for gradually-typed polymorphic
languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
To streamline such developments in Agda, we need support for a
step-indexed logic in Agda. This paper fills that need by showing how
to build a shallow embedding of a step-indexed logic in Agda, a system
that we abbreviate as SIL. To evaluate the utility of SIL, we present
a case study of proving semantic type safety of a simple cast
calculus, i.e., an internal language for the Gradually-Typed Lambda
Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. The outcome of that
study is that SIL streamlines the definition of logical relations and
is helpful in many of the proofs, but there is room for improvement
regarding the ergonomics of interactive proof development.
Limitations in Agda's inference algorithm sometimes leads to verbosity
in SIL proofs. It is worth noting that the authors have also proved
the Gradual Guarantee for the GTLC using SIL and the lesssons from
that experience confirm the findings in this paper.

TODO: ROAD MAP

\section{A Step-Indexed Logic (SIL)}

Figure~\ref{fig:SIL}

\begin{figure}
  \raggedright
Syntax
  \begin{syntax}
    \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
    \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
    \mid \overline{M} \in R\\
    \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
  \end{syntax}
Meaning of Propositions
\begin{align*}
  \sem{ P } \rho\; 0 &= \top \\
  \sem{ \top } \rho(suc \app n) &= \top \\
  \sem{ \bot } \rho (suc \app n) &= \bot \\
  \sem{ P \land Q } \rho (suc \app n) &=
    \sem{ P } \rho n \land \sem{ Q } \rho n \\
  \sem{ P \lor Q } \rho (suc \app n) &=
    \sem{ P } \rho n \lor \sem{ Q } \rho n \\
  \sem{ P \Rightarrow Q } \rho (suc \app n) &=
    \forall k \leq suc \app n,
    \sem{ P } \rho k \implies \sem{ Q } \rho k \\
  \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
  \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
  \sem{ \overline{M} \in R } \rho (suc \app n) &=
    \sem{ R } \rho (suc \app n) \app \overline{M}\\
  \sem{ \rhd P } \rho (suc \app n) &=
    \sem{ P } \rho n 
\end{align*}
Meaning of Relations
\begin{align*}
  \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
  \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
    \rho(r) (suc \app n) \app \overline{M} \\
  \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
  F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
  & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
  \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
    \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
\end{align*}
  \caption{Step-Indexed Logic (SIL)}
  \label{fig:SIL}
\end{figure}

\section{Case Study: Semantic Type Safety of a Cast Calculus}
